Low-Cost Data Deduplication for Virtual Machine Backup in Cloud Storage

### 2013-03-09 hotstorage13/arch.tex
correction

\section{ Key Ideas and Architecture}
\label{sect:arch}

@@

We consider deduplication in two levels. The first-level of data deduplication can be accomplished efficiently and 
inexpensively using a coarse-grain segment-level  dirty bit based detection in the OS level  to 
identify version difference from the previous snapshot to the current snapshot.  
Version-based detection has been used in the previous work~\cite{Clements2009,Vrable2009,TanIPDPS2011} and
we emphasize a segment-level coarse-grain setting to reduce the efforts in detection and maintaining dirty bits. 
Our experiment with Alibaba''s production dataset shows that over 70 percentage of 
duplicates can be detected using segment-based dirty bits while the segment size is as large as 2M bytes.  
This setting requires OS to keep the metadata for segment dirty bits and the amount of space for such segment 
dirty bits is negligible. In the second level of deduplication, content blocks of dirty segments at each VM 
are compared with the fingerprints of unique  blocks from the previous snapshots.

Our key ideas are summarized as follows.
\begin{itemize}

\item {\bf Separation of duplicate detection and data backup.}
%Request accumulation and partition-based deduplication.}
The second level duplicate detection requires a global comparison of 
content fingerprints of stored chunks with the new chunks scanned during 
the backup process. 
%Bloom filters and caching allows some of content 
%fingerprints to be stored on the cheap disks~\cite{bottleneck08}. Such an optimization is good 
%for inline deduplication where chunks scanned need to be determined to be duplicate or not instantly without 
%waiting while  it memory consumption is still significant in competing with stand virtual 
%machine activities.  
Our approach is to perform duplicate detection first before actual data backup.
That requires a pre-scanning of  data  blocks in dirty VM segments  and after that, real backup  for those non-duplicates
is conducted. That does incur an extra  round of local VM block reading while avoiding inline 
deduplication and leading to a much smaller resource requirement.  
During duplicate detection phase, detection requests can be accumulated on the disk 
for a period time  in a  partitioned manner.  Aggregated duplicate requests can be processed partition by partition. 
%We also accumulate the delete requests and perform them partition by partitions. 
Since each partition corresponds to a small portion of global content index, the cost to 
process requests within a partition is significantly smaller than processing all the requests globally.

\item {\bf Buffered data redistribution in parallel duplicate detection}.  
Let {\em global index} be the meta data containing the fingerprint values of unique blocks of snapshots
in the entire cluster and  the reference pointers which lead to the actual location of raw data.
%A duplicate detection request for a content block needs to be compared with the global index to determine if this
%block is non-duplicate. If it is a duplicate, return the corresponding reference pointer.
We conduct parallel processing of detection requests in all machines involved. 
A logical way to distribute detection requests is to partition based on fingerprint values of content blocks
in a  disjointed manner.
Initial data follows the VM distribution among machines and the detected results 
represented as  duplicate summary need to be collected following VM distribution. 
Therefore, there are two all-to-all data redistribution operations involved.
One is to map detection requests from VM-based distribution to fingerprint based distribution.  
One is to map duplicate summary from fingerprint-based distribution to VM based distribution.  
The redistributed data needs to be accumulated on the disk to minimize the use of memory.
To minimize the excessive disk seek cost, outgoing or incoming data exchange messages need to be buffered to bundle small messages.
Given there are $p\times q$ partitions where $p$ is the number of machines and $q$ is the number of fingerprint-based partitions
at each machine, the number of buffers is big and space per each buffer  is small given the memory constraint,
counteracting seek cost reduction.  
%Then there will be a large number of storage  IO operations involved, suffering the huge seek overhead.
We have designed an efficient data exchange and disk data buffering  scheme to address this.

@@

We assume a flat architecture in which  all $p$ machines that host VMs in a cluster can 
be used in parallel for deduplication. 
A small amount of local disk space and memory on each machine can be used 
to store global index and request accumulation.
The real backup storage can be either a distributed file system built on
this cluster  or use another  external storage system. 
%Our design is to minimize the usage of local memory and storage on each machine.

@@

%\subsection{Processing flow and resource usage}
The data flow of our multi-phase duplicate detection is depicted in Figure~\ref{fig:flow}. 
%let $v$ be the total number of virtual machines hosted in  a cluster,  
%we divide the backup into $k$  iterations. 
%\begin{itemize}
%\item
The first phase is that each machine independently reads  
%$v/k$  
virtual machine images that need a backup
and forms duplicate  detection requests. For those segments that have been modified since last backup indicated by 
dirty bits,  the system divides  each segment into a sequence of chunk blocks,  computes the meta 
information such as chunk fingerprints,  sends the metadata of a chunk block, and accumulates  into 
a partition bucket on the temporary disk storage. 
The partition mapping is using a hash function applied to the content fingerprint. 
Assuming all machines have a  homogeneous resource configuration, each machine is evenly  assigned with
$q$ parititons of content hash and accumulates corresponding requests on the disk. 
There are two options to assign buffers in each machine to buffer requests for each partition.
1) Each machine has  $p\times q$ send buffers corresponding to $p\times q$ partitions in the cluster
since a content block in a VM image of this machine can be sent to any of these partitions.
2) Each machine allocates $p$ send buffers correspond to send requests to $p$ machines and 
$p$ receive buffers to collect requests  from other machines.
Then the system copies requests in each $p$ disk buffers to  $q$ local disk buffers,
and outputs each disk buffer to one of the request partitions on the disk
when this disk buffer becomes full.  Option 2 is much more efficient than Option 1 because $2p+q$ is much smaller than
$p\times q$, as a result each buffer has a bigger size to accumulate data and results in 
much less disk seek overhead.

%We assume that the startup cost for network message sending is  much more less (e.g. an order of magnitude
%less) than from one machine to another 
%disk storage startup cost such as seek is much mu
%Note that we only accumulate requests with their meta data information.
%\item
The second phase is to load disk data and perform duplicate detection at each machine for  all accumulated 
requests one partition at a time.
% in the partitions.  
%The system maintains the index of all chunk fingerprints divided in the buckets. 
%The system loads the global index partition 
%and accumulated corresponding requests, and compare them to  identify the duplicated blocks.  
%Then we unload them, load the global index and accumulated requests for another partition. 
%This process is repeated until all buckets are processed.
At each iteration, as in-memory comparison between an index partition and request partition
is completed,  
duplicate information for segments of each VM is routed to  from fingerprint-based distribution  to VM-based distribution. 
The duplicate summary contains the block ID and  the reference pointer for each detected duplicate block.   
Each machine uses memory space of the request partition as a send buffer with no extra memory space required.
But it needs to allocate $p$ receive buffers to collect duplicate summary from other machines.
It also allocates $v$ disk buffers to copy duplicate summary from $p$ receive buffers and output to the local disk
when disk buffers are full.

@@

%\item 
The third phase is to perform real backup one VM at a time.
The system loads the duplicate summary of a VM obtained from Phase 2, 
reads  dirty segments of a VM, and output non-duplicate blocks to the final backup 
storage. Additionally, the global index on each machine is updated with the meta data of new chunk blocks. 
In this process, if a segment is not dirty, the system only needs to output the segment meta data 
such as reference pointers to the original segments previously saved in the backup storage.  
%If a segment is dirty,  the system  use the duplicate summary information from Step 2 for internal 
%duplicate blocks of this dirty segment and output the remaining non-duplicate blocks to the backup storage.
%\end{itemize}

@@

The above steps can be executed by each machine
using one thread to minimize the use of CPU, network bandwidth, and disk bandwidth.
The  disk storage usage on each machine 
is fairly small for  storing part of global index and
accumulating  duplicate detection requests that contain fingerprint information.   
We impose a limit for memory allocated at each machine node to accomplish low-cost deduplication.  
Memory usage for each phase of the above scheme is carefully controlled as follows.
\begin{itemize}
\item For Phase 1, the allocated memory for each machine is divided for 
1) a disk buffer to read dirty segments; 2) $2p$ send/receive buffers and $q$ disk buffers.

@@

\item For Phase 3 to  perform the real backup for each VM,
the allocated memory is divided for 1) a disk buffer to read dirty segments of a VM and   
write non-duplicate raw blocks to the  backup storage.
2) summary of duplicate blocks within dirty segments of  a VM.
\end{itemize}
%We find the memory for step 2 is most critical for buffering duplicate summary in hiding the latency. 
%Certainly most storage system can allow many IO requests issued in parallel overlapping the IO latency.  
%Here we assume the worst case and also minimize the IO bandwidth usages by issuing 
%one request at a time to use the minimal disk bandwidth resource.

{\bf Snapshot deletion.} Each VM will keep a limited number of  automatically-saved VM snapshots and thus
expired snapshots will be deleted unless its owner explicitly asks for its retention.
We adopt the idea of mark-and-sweep~\cite{MarkSweep}. 
A block or a segment can be deleted if its reference count is zero.
%As we discussed in Section~\ref{sect:data}, the reference counter is kept in the global index, partitioned among machines. 
To delete useless  blocks or segments periodically, we read the meta data  of live 
snapshots and compute the reference count of all blocks and segments  in parallel
while following the fingerprint-based partitioning.
Similar to the multi-phase duplicate detection process, reference counting is conducted in multi-phases. 
Phase  1 is to read  the metadata of snapshots such as segment and block
recipes and  to accumulate  reference count requests in different machines  in the fingerprint based distribution.
Phase   2 is to count references within each partition and detect those records with zero 
reference. 
%Phase 3 is that every VM gathers its z confirmed deletion requests for its content and sends them to the backup storage.
The backup  data repository logs deletion instructions,  and will periodically perform a compaction operation when 
its deletion log is too big. 
%Due to the paper constraint, we will not discuss the deletion operations in details.

@@

%Since we maintain a single copy of each block information in the bucket index, and there is at most one bucket worker operating the bucket index at one time, it it safe to do the simple reference counting rather than mark-and-sweep.

>>>>

\section{System Design}
\label{sect:arch}

@@

We consider deduplication in two levels. The first-level of data deduplication can be accomplished efficiently and 
inexpensively using a coarse-grain segment-level  dirty bit based detection in the OS level  to 
identify version difference from the previous snapshot to the current snapshot.  
Version-based detection has been used in the previous work~\cite{Clements2009,Vrable2009,TanIPDPS2011} and
we emphasize a segment-level coarse-grain setting to reduce the efforts in detection and maintaining dirty bits. 
Our experiment with Alibaba's production dataset shows that over 70 percentage of 
duplicates can be detected using segment-based dirty bits when the segment size is as large as 2M bytes.  
This setting requires OS to keep the metadata for segment dirty bits and the amount of space for such segment 
dirty bits is negligible. In the second level of deduplication, content blocks of dirty segments at each VM 
are compared with the fingerprints of unique  blocks from the previous snapshots.

Our key strategies are explained as follows.
\begin{itemize}

\item {\bf Separation of duplicate detection and data backup.}
%Request accumulation and partition-based deduplication.}
The second level duplicate detection requires a global comparison of 
content fingerprints.
% of stored chunks with the new chunks scanned during the backup process. 
%Bloom filters and caching allows some of content 
%fingerprints to be stored on the cheap disks~\cite{bottleneck08}. Such an optimization is good 
%for inline deduplication where chunks scanned need to be determined to be duplicate or not instantly without 
%waiting while  it memory consumption is still significant in competing with stand virtual 
%machine activities.  
Our approach is to perform duplicate detection first before actual data backup.
That requires a pre-scanning of  data  blocks in dirty VM segments  and after that, real backup  for those non-duplicates
is conducted. That does incur an extra  round of local VM block reading while avoiding inline 
deduplication and leading to a much smaller resource requirement.  
During VM reading, detection requests can be accumulated on the disk 
for a period time  in a  partitioned manner.  Aggregated duplicate requests can be processed partition by partition. 
%We also accumulate the delete requests and perform them partition by partitions. 
Since each partition corresponds to a small portion of global content index, the cost to 
process detection requests within a partition is significantly smaller than processing all the requests globally.

\item {\bf Buffered data redistribution in parallel duplicate detection}.  
Let {\em global index} be the meta data containing the fingerprint values of unique blocks of snapshots
in the entire cluster and  the reference pointers which lead to the actual location of raw data.
%A duplicate detection request for a content block needs to be compared with the global index to determine if this
%block is non-duplicate. If it is a duplicate, return the corresponding reference pointer.
We conduct parallel processing of detection requests in all machines involved. 
A logical way to distribute detection requests is to partition based on fingerprint values of content blocks
in a  disjointed manner.
Initial VM data reading follows the VM distribution among machines and the detected duplicate summary results 
need to be collected following the VM distribution. 
Therefore, there are two all-to-all data redistribution operations involved.
One is to map detection requests from VM-based distribution to fingerprint based distribution.  
One is to map duplicate summary from fingerprint-based distribution to VM based distribution.  
The redistributed data needs to be accumulated on the disk to minimize the use of memory.
To minimize the excessive disk seek cost, outgoing or incoming data exchange messages need to be buffered to bundle small messages.
Given there are $p\times q$ partitions where $p$ is the number of machines and $q$ is the number of fingerprint-based partitions
at each machine, space per each buffer  is small under the memory constraint for large $p$ or $q$ values.
This counteracts the effort of seek cost reduction.  
%Then there will be a large number of storage  IO operations involved, suffering the huge seek overhead.
We have designed an efficient data exchange and disk data buffering  scheme to address this.

@@

We assume a flat architecture in which  all $p$ machines that host VMs in a cluster can 
be used in parallel for deduplication. 
A small amount of local disk space and memory on each machine can be used 
to store global index and temporary data. 
The real backup storage can be either a distributed file system built on
this cluster  or use another  external storage system. 
%Our design is to minimize the usage of local memory and storage on each machine.

@@

%\subsection{Processing flow and resource usage}
The data flow of our multi-phase duplicate detection is depicted in Figure~\ref{fig:flow}. 
%let $v$ be the total number of virtual machines hosted in  a cluster,  
%we divide the backup into $k$  iterations. 
%\begin{itemize}
%\item
In the first phase, each machine independently reads  
%$v/k$  
VM images that need a backup
and forms duplicate  detection requests. For those segments that have been modified since last backup indicated by 
dirty bits,  the system divides  each segment into a sequence of chunk blocks,  computes the meta 
information such as chunk fingerprints,  sends a request to a proper machine, accumulates  
received requests into a partition on the local temporary disk storage. 
The partition mapping is using a hash function applied to the content fingerprint. 
Assuming all machines have a  homogeneous resource configuration, each machine is evenly  assigned with
$q$ parititons of content hash and accumulates corresponding requests on the disk. 
There are two options to assign buffers in each machine to buffer requests for each partition.
1) Each machine has  $p\times q$ send buffers corresponding to $p\times q$ partitions in the cluster
since a content block in a VM image of this machine can be sent to any of these partitions.
2) Each machine allocates $p$ send buffers to deliver requests to $p$ machines and 
$p$ receive buffers to collect requests  from other machines.
Then the system copies requests from each of $p$ receive buffers to  $q$ local disk buffers,
and outputs each disk buffer to one of the request partitions on the disk
when this disk buffer becomes full.  Option 2 is much more efficient than Option 1 because $2p+q$ is much smaller than
$p\times q$. As a result, each buffer has a bigger size to accumulate requests and that means
less disk seek overhead.

%We assume that the startup cost for network message sending is  much more less (e.g. an order of magnitude
%less) than from one machine to another 
%disk storage startup cost such as seek is much mu
%Note that we only accumulate requests with their meta data information.
%\item
The second phase is to load disk data and perform fingerprint comparison at each machine for  all accumulated 
requests one partition at a time.
% in the partitions.  
%The system maintains the index of all chunk fingerprints divided in the buckets. 
%The system loads the global index partition 
%and accumulated corresponding requests, and compare them to  identify the duplicated blocks.  
%Then we unload them, load the global index and accumulated requests for another partition. 
%This process is repeated until all buckets are processed.
At each iteration, once in-memory comparison between an index partition and request partition is completed,  
duplicate summary information for segments of each VM is routed from the fingerprint-based distribution  to the
VM-based distribution.  The duplicate summary contains the block ID and  the reference pointer for each detected duplicate block.   
Each machine uses memory space of the request partition as a send buffer with no extra memory space required.
But it needs to allocate $p$ receive buffers to collect duplicate summary from other machines.
It also allocates $v$ disk buffers to copy duplicate summary from $p$ receive buffers and output to the local disk
when disk buffers are full.

@@

%\item 
The third phase is to perform real backup one VM at a time.
The system loads the duplicate summary of a VM obtained from Phase 2, 
reads  dirty segments of a VM, and output non-duplicate blocks to the final backup 
storage. Additionally, the global index on each machine is updated with the meta data of new chunk blocks. 
In this process, if a segment is not dirty, the system only needs to output the segment meta data 
such as a reference pointer to the original segment previously saved in the backup storage.  
%If a segment is dirty,  the system  use the duplicate summary information from Step 2 for internal 
%duplicate blocks of this dirty segment and output the remaining non-duplicate blocks to the backup storage.
%\end{itemize}

@@

The above steps can be executed by each machine
using one thread to minimize the use of CPU, network bandwidth, and disk bandwidth.
The  disk storage usage on each machine 
is fairly small for  storing part of global index and
accumulating  duplicate detection requests that contain fingerprint information.   
We impose a memory limit $M$ allocated for each phase of processing at each machine.
The usage of $M$ is controlled as follows and the detailed distribution is optimized based on the relative
ratio between the cross-machine network  startup cost and disk access startup cost.
\begin{itemize}
\item For Phase 1, the allocated memory for each machine is divided for 
1) a disk buffer to read dirty segments; 2) $2p$ send/receive buffers and $q$ disk buffers.

@@

\item For Phase 3, the allocated memory is divided for 1) an I/O buffer to read dirty segments of a VM and   
write non-duplicate content blocks to the  backup storage.
2) summary of duplicate blocks within dirty segments of  a VM.
\end{itemize}
%We find the memory for step 2 is most critical for buffering duplicate summary in hiding the latency. 
%Certainly most storage system can allow many IO requests issued in parallel overlapping the IO latency.  
%Here we assume the worst case and also minimize the IO bandwidth usages by issuing 
%one request at a time to use the minimal disk bandwidth resource.

{\bf Snapshot deletion.} Each VM will keep a limited number of  automatically-saved VM snapshots and thus
expired snapshots will be deleted unless its owner explicitly asks for its retention.
We adopt the idea of mark-and-sweep~\cite{Guo2011}. 
A block or a segment can be deleted if its reference count is zero.
%As we discussed in Section~\ref{sect:data}, the reference counter is kept in the global index, partitioned among machines. 
To delete useless  blocks or segments periodically, we read the meta data  of live 
snapshots and compute the reference count of all blocks and segments  in parallel
while following the fingerprint-based partitioning.
Similar to the multi-phase duplicate detection process, reference counting is conducted in multi-phases. 
Phase  1 is to read  the metadata of snapshots such as segment and block
recipes and  to accumulate  reference count requests in different machines  in the fingerprint based distribution.
Phase   2 is to count references within each partition and detect those records with zero 
reference. 
%Phase 3 is that every VM gathers its z confirmed deletion requests for its content and sends them to the backup storage.
The backup  data repository logs deletion instructions,  and will periodically perform a compaction operation when 
its deletion log is too big. 
%Due to the paper constraint, we will not discuss the deletion operations in details.

@@

%Since we maintain a single copy of each block information in the bucket index, and there is at most one bucket worker operating the bucket index at one time, it it safe to do the simple referencec ounting rather than mark-and-sweep.
